---
title: "Predicting the Types of Biceps Curls"
author: "Warren San Luis"
date: "September 23, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, echo=FALSE, message=FALSE}
library(data.table)
library(caret)
library(rpart.plot)
```

## Executive Summary

The goal of this report is to detail an analysis of participants performing Unilateral Dumbbell Bicepts Curls. Using accelerometers, we look to predict the way the curls are performed.  There are five different ways (classes) that were examined.

```{r get data, echo=FALSE}
setwd("C:\\Users\\Warren\\Desktop\\Coursework\\Coursera\\Practical Machine Learning\\Course Project")

if (!file.exists("modeling.csv")) {
  download.file(url = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv",
                destfile = "modeling.csv")
}
  
if (!file.exists("holdout.csv")) {
  download.file(url = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv",
                destfile = "holdout.csv")
}

# READ IN CSVS
modeling <- read.csv("modeling.csv")
holdout <- read.csv("holdout.csv")
```

## Examining the Data
The data was examined for data quality. After examination, it was found that the columns that contained the following terms had mostly blank or NAs: kurtosis, skewness, max, min, amplitude, var, avg, stddev. These were removed from the dataset.

The main dataset analyzed contained 19,622 observations. A separate holdout dataset of 20 observations was used as an out-of-sample set and was not used in modeling.

```{r remove cols, echo=FALSE}
# Removing certain columns due to bad or lack of data in certain columns
model_cols <- as.vector(names(modeling))
include_flag <- !grepl("kurtosis|skewness|max|min|amplitude|var|avg|stddev", model_cols)
modeling2 <- modeling[, include_flag][,-c(1:7)]
```

## Fitting the Models

The main dataset was split into a training (70% of data) and test set (30% of data). Each model was first fit to the training set and checked on the test set. The models used were a decision tree with length = 50, a linear discriminant analysis, and a quadratic discriminant analysis.

Reasons for selecting/not selecting models to investigate:  

* Usage of a tree allowed for more flexibility in predictions.
* Linear discriminant analysis was used to investigate linear decision boundaries.
* Quadtratic discriminant analysis was used to investigate more non-linear decision boundaries.
* Random forests were not investigated due to long runtime.

```{r traintest, echo=FALSE}
# Removing certain columns due to bad or lack of data in certain columns
model_cols <- as.vector(names(modeling))
include_flag <- !grepl("kurtosis|skewness|max|min|amplitude|var|avg|stddev", model_cols)
modeling2 <- modeling[, include_flag][,-c(1:7)]

# Split modeling data into training and test sets
set.seed(62351)
inTrain <- createDataPartition(modeling2$classe, times = 1, p = 0.7)[[1]]

training <- modeling2[inTrain,]
testing <- modeling2[-inTrain,]
```

The results on the test set are shown in the **Appendix**.

The decision tree and quadratic discriminant analysis had much stronger results than the linear discriminant analysis. The decision tree and quadratic discriminant analysis models are highly flexible and can be subject to overfitting. To help account for the flexible model fitting, we checked these models against the testing set that was set aside. The accuracy, sensitivity, and specificity were all high except for the linear discriminant analysis. The decision tree results were slightly more favorable than the quadratic discriminant analysis where the error rates were both near 10%.
  
## Final Results
When applying the decision tree results to the 20 observations that were held out from the modeling, 18 of the 20 observations were correctly predicted which implies an out-of-sample error of 10%. The final predictions on these 20 observations by model type are shown in the **Appendix**.

```{r fits, echo=FALSE, message=FALSE}
# CREATE DECISION TREE AND CONFUSION MATRIX

# Fit tree
fit_dectree <- train(classe ~ ., data = training, method = "rpart", tuneLength = 50)

# Create predictions in testing
pred_dectree <- predict(fit_dectree, newdata = testing)

# Attach predictions to modeling dataset
testing_dectree <- cbind(testing, pred_dectree)

# CREATE LINEAR DISCRIMINANT ANALYSIS

# Fit model
fit_lda <- train(classe ~ ., data = training, method = "lda")

# Create predictions in testing
pred_lda <- predict(fit_lda, newdata = testing)

# Attach predictions to testing dataset
testing_lda <- cbind(testing, pred_lda)

# CREATE QUADRATIC DISCRIMINANT ANALYSIS

# Fit model
fit_qda <- train(classe ~ ., data = training, method = "qda")

# Create predictions in testing
pred_qda <- predict(fit_qda, newdata = testing)

# Attach predictions to modeling dataset
testing_qda <- cbind(testing, pred_qda)
```

## Appendix

**FIGURE A - DECISION TREE (LENGTH = 50)**
```{r dectree, echo=FALSE}
# Create confusion matrix on testing data
confusionMatrix(testing_dectree$pred_dectree, testing_dectree$classe)
```

**FIGURE B - LINEAR DISCRIMINANT ANALYSIS**
```{r lda, echo=FALSE}
# Create confusion matrix on testing data
confusionMatrix(testing_lda$pred_lda, testing_lda$classe)
```

**FIGURE C - QUADRATIC DISCRIMINANT ANALYSIS**
```{r qda, echo=FALSE}
# Create confusion matrix on testing data
confusionMatrix(testing_qda$pred_qda, testing_qda$classe)
```

**FIGURE D - MODEL RESULTS ON HOLDOUT (20 OBSERVATIONS)**
```{R holdout, echo=FALSE}

holdout2 <- holdout[,include_flag][,-c(1:7)]

# Decision tree (length = 50)
hold_dectree <- predict(fit_dectree, newdata = holdout2)

# Linear discriminant analysis
hold_lda <- predict(fit_lda, newdata = holdout2)

# Quadratic discriminant analysis
hold_qda <- predict(fit_qda, newdata = holdout2)

# Put all models together
holdout_final <- data.table(cbind(holdout2, hold_dectree, hold_lda, hold_qda))

# Print results
holdout_final2 <- holdout_final[,.(problem_id,
                                   hold_dectree,
                                   hold_qda,
                                   hold_lda)]
holdout_final2
```